Too Beautiful to be Fake: Attractive Faces are Less Likely to be Judged as Artificially Generated

### 2024-06-18 submission/new_manuscript/manuscript.qmd
updated revision doc and manuscript doc

AI-generated content, in particular realistic images, is becoming commonplace and carries important risks for misinformation and black-mailing [@viola2023designed], emphasizing the need to understand the different components that come into play in the formation of reality beliefs.
This exploratory study primarily aims at investigating the effect of facial attractiveness on simulation monitoring, i.e., on the beliefs that an image is real or artificially generated. Based on the *affective reality theory*
[outlined in @makowski2018cognitive; @makowski2019phenomenal, see @makowski2023affective for a summary],
which suggests that salient and emotional stimuli are perceived to be more real (up to a point of reversal after which beliefs of fiction becomes used an emotion regulation strategy), we hypothesize a quadratic relationship between perceived realness and attractiveness: faces rated as highly attractive or unattractive will more likely be believed to be real. We expect a similar relationship with trustworthiness ratings given its well-established link with attractiveness [@liefooghe2022faces; @bartosik2021you; @garrido2017kdef; @little2011facial], and a positive relationship with familiarity (as more familiar faces would appear as more salient, self-relevant and anchored in reality). Additionally, we will further explore the link shared by dispositional traits, such as personality and attitude towards AI, with simulation monitoring tendencies. Importantly, this study does investigate the discriminative accuracy between "true" photos and "true" artificially-generated images (which we consider more a technological issue than a psychological one), focusing on the beliefs that a stimulus is real or fake, independently of its true nature.

@@

All the material (preregistration, experiment demo, experiment code, raw data, analysis script with complementary figures and analyses, etc.) is available at [**MASKED FOR REVIEW**]
<!-- [**https://github.com/RealityBending/FakeFace**](https://github.com/RealityBending/FakeFace){.uri}. -->

### Ethics Statement

This study was approved by the 
<!-- NTU Institutional Review Board (NTU IRB-2022-187)  -->
[**IRB NUMBER**]
and all procedures performed were in accordance with the ethical standards of the institutional board and with the 1964 Helsinki Declaration. All participants provided their informed consent prior to participation and were incentivized after completing the study.

### Procedure

In the first part of the study, participants answered a series of personality questionnaires presented in the order below. These include the *Mini-IPIP6*
<!-- (24 items) [@sibley2011]  -->
[24 items, @sibley2011]
measuring 6 personality traits, the *SIAS-6* and the *SPS-6*
<!-- (6 items each) [@peters2012development] -->
[6 items each, @peters2012development]
assessing social anxiety levels, 5 items we devised pertaining to expectations about AI-generated image technology ("I think current Artificial Intelligence algorithms can generate very realistic images"), of which we mixed with 5 items from the general attitudes towards AI scale to lower the former's saliency and the possibility of it priming the subjects about the task,
<!-- (*GAAIS*) [@schepman2020initial]. -->
[*GAAIS*, @schepman2020initial]
the *FFNI-BF*
<!-- (30 items) [@jauk2022validation] -->
[30 items, @jauk2022validation]
measuring 9 facets of narcissism; the *R-GPTS*
<!-- (18 items) [@freeman2021revised] -->
[18 items, @freeman2021revised]
measuring 2 dimensions related to paranoid thinking; and the *IUS-12*
<!-- (12 items) [@carleton2007fearing] -->
[12 items, @carleton2007fearing]
measuring intolerance to uncertainty. Self-rated attractiveness was also assessed using 2 items - one measuring general attractiveness
<!-- ("How attractive would you say you are?") [@marcinkowska2021self] -->
["How attractive would you say you are?", @marcinkowska2021self]
and the other measuring physical attractiveness
<!-- ("How would you rate your own physical attractiveness relative to the average") [@spielmann2020predictive] -->
["How would you rate your own physical attractiveness relative to the average", @spielmann2020predictive].
3 attention check questions were also embedded in the surveys.

@@

### Participants

Although the main part of the study relied on within-subject design (with 109 trials per participant), we also planned to do between-participants analyses, thus aiming at collecting a larger sample than traditionally used in experimental psychology (with budget availability as the main constraint). One hundred and fifty participants were recruited via *Prolific*, a crowd-sourcing platform recognized for providing high quality data [@peer2022; douglas2023data]. The only inclusion criterion was a fluent proficiency in English to ensure that the experiment instructions would be well-understood. Participants were incentivised with a reward of about \textsterling 7.5 for completing the study, which took about 45 minutes to finish. Demographic variables (age, gender, sexual orientation, education and ethnicity) were self-reported on a voluntary basis.

@@

### Data Analysis

The real-fake ratings (measured originally with a [-1, 1] analog scale) were converted into two scores, corresponding to two conceptually distinct mechanisms: the dichotomous *belief* (real or fake, based on the sign of the rating) and the *confidence* (the rating's absolute value) associated with that belief. The former was analyzed using logistic mixed models, which modelled the probability of assigning a face to the real (\>= 0) as opposed to fake (\< 0). The latter, as well as the other face ratings (attractiveness, beauty, trustworthiness and familiarity), was modelled using mixed beta regressions (suited for outcome variables expressed in percentages). The models included the participants and stimuli as random factors.

@@

### Determinants of Simulation Monitoring

Attractiveness had a significant positive and linear relationship ($R^2_{marginal}$ = 2.0%) with the belief that a stimulus was real ($\beta_{poly1} = 16.57$, $95\%~CI = [7.33, 25.82]$, $z = 3.51$, $p < .001$) for males, and a quadratic relationship for females ($\beta_{poly2} = 7.82$, $95\%~CI = [1.81, 13.84]$, $z = 2.55$, $p = .011$), with both non-attractive and attractive faces being judged as more real. Attractiveness was also found to have a significant positive and quadratic relationship with confidence in judging faces both as real ($\beta_{poly2} = 4.30$, $95\%~CI = [0.97, 7.64]$, $z = 2.53$, $p = .011$) and as fake ($\beta_{poly2} = 5.23$, $95\%~CI = [0. 86, 9.60]$, $z = 2.35$, $p = .019$) for females. For males, however, a significant negative and quadratic relationship was found between attractiveness ratings and belief confidence only for faces judged as fake ($\beta_{poly2} = -9.92$, $95\%~CI = [-18.99, -0.86]$, $z = -2.15$, $p = .032$). There was no interaction with reported self-attractiveness.

@@

### Inter-Individual Correlates of Simulation Monitoring

The models including the personality traits suggested that *Honesty-Humility* had a significant negative relationship with the confidence associated with real as well as fake judgements ($\beta_{real} = -1.62$, $95\%~CI = [-2.55, -0.70]$, $z = -3.43$, $p < .001$; $\beta_{fake} = -1.16$, $95\%~CI = [-2.09, -0.23]$, $z = -2.45$, $p = 0.014$).

@@

The datasets generated and/or analysed during the current study are available in the GitHub repository 
<!-- https://github.com/RealityBending/FakeFace -->

>>>>

AI-generated content, in particular realistic images, is becoming commonplace and carries important risks for misinformation and black-mailing [@viola2023designed], emphasizing the need to understand the different components that come into play in the formation of reality beliefs.
This exploratory study primarily aims at investigating the effect of facial attractiveness on simulation monitoring, i.e., on the beliefs that an image is real or artificially generated. Based on the *affective reality theory* [@makowski2023affective], which suggests that salient and emotional stimuli are perceived to be more real (up to a point of reversal after which beliefs of fiction becomes used an emotion regulation strategy), we hypothesize a quadratic relationship between perceived realness and attractiveness: faces rated as highly attractive or unattractive will more likely be believed to be real. We expect a similar relationship with trustworthiness ratings given its well-established link with attractiveness [@liefooghe2022faces; @bartosik2021you; @garrido2017kdef; @little2011facial], and a positive relationship with familiarity (as more familiar faces would appear as more salient, self-relevant and anchored in reality). Additionally, we will further explore the link shared by dispositional traits, such as personality and attitude towards AI, with simulation monitoring tendencies. Importantly, this study does investigate the discriminative accuracy between "true" photos and "true" artificially-generated images (which we consider more a technological issue than a psychological one), focusing on the beliefs that a stimulus is real or fake, independently of its true nature.

@@

All the material (preregistration, experiment demo, experiment code, raw data, analysis script with complementary figures and analyses, etc.) is available at [**https://github.com/RealityBending/FakeFace**](https://github.com/RealityBending/FakeFace){.uri}.

**Ethics Statement.** 
This study was approved by the NTU Institutional Review Board (NTU IRB-2022-187)
<!-- [**IRB NUMBER**] -->
and all procedures performed were in accordance with the ethical standards of the institutional board and with the 1964 Helsinki Declaration. All participants provided their informed consent prior to participation and were incentivized after completing the study.

**Procedure.** 
In the first part of the study, participants answered a series of personality questionnaires presented in the order below. These include the *Mini-IPIP6*
<!-- (24 items) [@sibley2011]  -->
[24 items, @sibley2011]
measuring 6 personality traits, the *SIAS-6* and the *SPS-6*
<!-- (6 items each) [@peters2012development] -->
[6 items each, @peters2012development]
assessing social anxiety levels, 5 items we devised pertaining to expectations about AI-generated image technology ("I think current Artificial Intelligence algorithms can generate very realistic images"), of which we mixed with 5 items from the general attitudes towards AI scale to lower the former's saliency and the possibility of it priming the subjects about the task,
<!-- (*GAAIS*) [@schepman2020initial]. -->
[*GAAIS*, @schepman2020initial]
the *FFNI-BF*
<!-- (30 items) [@jauk2022validation] -->
[30 items, @jauk2022validation]
measuring 9 facets of narcissism; the *R-GPTS*
<!-- (18 items) [@freeman2021revised] -->
[18 items, @freeman2021revised]
measuring 2 dimensions related to paranoid thinking; and the *IUS-12*
<!-- (12 items) [@carleton2007fearing] -->
[12 items, @carleton2007fearing]
measuring intolerance to uncertainty. Self-rated attractiveness was also assessed using 2 items - one measuring general attractiveness
<!-- ("How attractive would you say you are?") [@marcinkowska2021self] -->
["How attractive would you say you are?", @marcinkowska2021self]
and the other measuring physical attractiveness
<!-- ("How would you rate your own physical attractiveness relative to the average") [@spielmann2020predictive] -->
["How would you rate your own physical attractiveness relative to the average", @spielmann2020predictive].
3 attention check questions were also embedded in the surveys.

@@

**Participants.**
Although the main part of the study relied on within-subject design (with 109 trials per participant), we also planned to do between-participants analyses, thus aiming at collecting a larger sample than traditionally used in experimental psychology (with budget availability as the main constraint). One hundred and fifty participants were recruited via *Prolific*, a crowd-sourcing platform recognized for providing high quality data [@peer2022; douglas2023data]. The only inclusion criterion was a fluent proficiency in English to ensure that the experiment instructions would be well-understood. Participants were incentivised with a reward of about \textsterling 7.5 for completing the study, which took about 45 minutes to finish. Demographic variables (age, gender, sexual orientation, education and ethnicity) were self-reported on a voluntary basis.

@@

**Data Analysis.**
The real-fake ratings (measured originally with a [-1, 1] analog scale) were converted into two scores, corresponding to two conceptually distinct mechanisms: the dichotomous *belief* (real or fake, based on the sign of the rating) and the *confidence* (the rating's absolute value) associated with that belief. The former was analyzed using logistic mixed models, which modelled the probability of assigning a face to the real (\>= 0) as opposed to fake (\< 0). The latter, as well as the other face ratings (attractiveness, beauty, trustworthiness and familiarity), was modelled using mixed beta regressions (suited for outcome variables expressed in percentages). The models included the participants and stimuli as random factors.

@@

**Determinants of Simulation Monitoring.**
Attractiveness had a significant positive and linear relationship ($R^2_{marginal}$ = 2.0%) with the belief that a stimulus was real ($\beta_{poly1} = 16.57$, $95\%~CI = [7.33, 25.82]$, $z = 3.51$, $p < .001$) for males, and a quadratic relationship for females ($\beta_{poly2} = 7.82$, $95\%~CI = [1.81, 13.84]$, $z = 2.55$, $p = .011$), with both non-attractive and attractive faces being judged as more real. Attractiveness was also found to have a significant positive and quadratic relationship with confidence in judging faces both as real ($\beta_{poly2} = 4.30$, $95\%~CI = [0.97, 7.64]$, $z = 2.53$, $p = .011$) and as fake ($\beta_{poly2} = 5.23$, $95\%~CI = [0. 86, 9.60]$, $z = 2.35$, $p = .019$) for females. For males, however, a significant negative and quadratic relationship was found between attractiveness ratings and belief confidence only for faces judged as fake ($\beta_{poly2} = -9.92$, $95\%~CI = [-18.99, -0.86]$, $z = -2.15$, $p = .032$). There was no interaction with reported self-attractiveness.

@@

**Inter-Individual Correlates of Simulation Monitoring.**
The models including the personality traits suggested that *Honesty-Humility* had a significant negative relationship with the confidence associated with real as well as fake judgements ($\beta_{real} = -1.62$, $95\%~CI = [-2.55, -0.70]$, $z = -3.43$, $p < .001$; $\beta_{fake} = -1.16$, $95\%~CI = [-2.09, -0.23]$, $z = -2.45$, $p = 0.014$).

@@

The datasets generated and/or analysed during the current study are available in the GitHub repository 
https://github.com/RealityBending/FakeFace
